Follow the Flow of Rain and Paint: \break An Output Sensitive Contour Tree Algorithm

### 2014-06-28 paint-contour-tree.tex
small changes to section 3

We begin by associating a color with each maximum. If the set of maxima is $S$,
this is simply an arbitrary bijection $\chi:S \mapsto [|S|]$. Imagine there being a large
can of paint of color $c(y)$ at maxima $y$. We will spill different paint from each maximum and watch it flow down.
This is analogous to the raining in the previous section, but paint is a much more viscous liquid.
\emph{So paint only flows down edges, and it does not color the interior of faces.} Furthermore, paints
do not mix, so eventually, every vertex and edge of $\MM$ gets a unique color.
%
\begin{definition} \label{def:paint} A  \emph{painting} of $\MM$ is a map $\chi:V \cup E \mapsto [|S|]$
with the following properties. (Remember that $G(\MM) = (V,E)$.)
\begin{asparaitem}
	\item The restriction $c:S \mapsto [|S|]$ is a bijection.
	\item Consider an edge $e$. There exists a descending path from some maximum $x$ to $e$
	consisting of edges in $E$, such that all edges and vertices along this path have the same color. (The maximum $x$
	is well-defined to be the unique maximum with the same color as $e$.)
\end{asparaitem}
\end{definition}
%
Note that paintings are not unique. It is straightforward to construct a painting of $\MM$ in $O(n)$ time, using a descending BFS from each maximum.

@@

{\bf The binomial heaps $\touch(c)$:} For each color $c$, we store a binomial
heap $T(c)$ of non-extremal critical points.

\begin{definition} \label{def:color-set} Fix a painting $\chi$, and a color $c$.
A critical point $x$ is \emph{touched by $c$} if $x$ is incident to a $c$-colored
increasing edge. The set of non-extremal critical points touched by $c$ is denoted $\touch(c)$.
For $x$, $\col(x)$ is the set of colors that touch $x$.
\end{definition}
%
We will store each set $\touch(c)$ in a binomial heap, keyed by the heights. 
Abusing notation, $\touch(c)$ refers
both to the set and the data structure used to store it.

@@

\medskip
\noindent
{\bf The stack $\st$:} This consists of non-extremal critical points with the following structure.
Each point $x \in \st$ has an associated subset of $\col(x)$, denoted $\mcol(x)$.
Both $\mcol(x)$ and its complement are stored as hash table. So lookups, inserts, and deletes
are in these sets are all constant time operations. The stack is guaranteed to satisfy 
the following invariants.
\begin{asparaitem}
	\item For every $x \in \st$: For every $c \in \mcol(x)$, $x$ is the highest element
	in $T(c)$. Furthermore, $c = \rep(c)$.
	\item Consider $x, y \in \st$ such that $y$ was pushed on $x$. There exists $c \in \col(x) \setminus
	\mcol(x)$ such that $c \in \col(y)$.
\end{asparaitem}

@@

\smallskip
\begin{compactenum}
	\item If $K$ is empty, pick arbitrary $c$. Let $x$ be the highest point in $T(c)$.
	Push $x$ onto stack and set $\mcol(x) = \{c\}$.
	\item Let $h$ be head of $\st$.
	\item While $\col(h) \setminus \mcol(h)$ non-empty:
	\begin{compactenum}
		\item Pick some $c \in \col(h) \setminus \mcol(h)$. Replace it by $c' = \rep(c)$.
		\item Obtain highest point $x$ in $T(c')$. 
		\item If $x = h$, put $c'$ in $\mcol(h)$. Otherwise ($x \neq h$), push $x$ onto stack
		and set $\mcol(x) = \{c'\}$.
		\item Let $h$ be new head of $\st$.
	\end{compactenum}
\end{compactenum}
\end{minipage}}

@@

\smallskip
\begin{compactenum}
	\item Run $\update(\st)$. Pop $\st$ to get $h$.
	\item Suppose $\mcol(h) = \{c\}$ (so $h$ is split). 
	\begin{compactenum}
		\item Connect $h$ (in $\reeb(\MM)$) to $\h(C)$ and the unique minimum corresponding to split $h$. 
		\item Delete $h$ from $T(c)$, set $\h(C) = h$.
		\item End procedure.
	\end{compactenum}
	\item Let $\mcol(h) = \{c_1, c_2\}$ (so $h$ is merge).
	\item Connect $h$ in $\reeb(\MM)$ to $\h(c_1)$ and $\h(c_2)$.
	\item Delete all copies of $h$ from $T(c_1)$ and $T(c_2)$.
	\item Perform union of colors $c_1$ and $c_2$ (denote merged color as $c$). Merge heaps
	to get $T(c) = T(c_1) \cup T(c_2)$.
	\item Set $\h(c) = h$.	
\end{compactenum}
\end{minipage}}

>>>>

We begin by associating a color with each maximum. If the set of maxima is $S$,
this is simply an arbitrary bijection $\chi:S \mapsto [|S|]$. Imagine there being a large
can of paint of color $\chi(y)$ at maxima $y$. We will spill different paint from each maximum and watch it flow down.
This is analogous to the raining in the previous section, but paint is a much more viscous liquid.
\emph{So paint only flows down edges, and it does not color the interior of faces.} Furthermore, paints
do not mix, so eventually, every vertex and edge of $\MM$ gets a unique color.
%
\begin{definition} \label{def:paint} A  \emph{painting} of $\MM$ is a map $\chi:S \cup E \mapsto [|S|]$
with the following properties. (Remember that $G(\MM) = (V,E)$.)
\begin{asparaitem}
	\item The restriction $\chi:S \mapsto [|S|]$ is a bijection.
	\item Consider an edge $e$. There exists a descending path from some maximum $x$ to $e$
	consisting of edges in $E$, such that all edges along this path have the same color as $x$. (The maximum $x$
	is well-defined to be the unique maximum with the same color as $e$.)
\end{asparaitem}
\end{definition}
%
Note that paintings are not unique. It is straightforward to construct a painting of $\MM$ in $O(n)$ time, using a descending BFS from each maximum that does not explore previously colored edges.

@@

\begin{definition} \label{def:color-set} Fix a painting $\chi$, and a color $c$.
A critical point $x$ is \emph{touched by $c$} if $x$ is incident to a $c$-colored
increasing edge. The set of non-extremal critical points touched by $c$ is denoted $\touch(c)$.
For $x$, $\col(x)$ is the set of colors that touch $x$.
\end{definition}
%
%We will store each set $\touch(c)$ in a binomial heap, keyed by the heights. 
Abusing notation, $\touch(c)$ refers
both to the set and the data structure used to store it.

\medskip
\noindent
{\bf The binomial heaps $\touch(c)$:} For each color $c$, we store a binomial
heap $T(c)$ of non-extremal critical points, keyed by vertex heights.

@@

\medskip
\noindent
{\bf The stack $\st$:} This consists of non-extremal critical points with the following structure.
Each point $x \in \st$ has an associated subset of $\col(x)$, denoted $\mcol(x)$.
Both $\mcol(x)$ and its complement are stored as hash table. So lookups, inserts, and deletes
are in these sets are all constant time operations. The stack is guaranteed to satisfy 
the following invariants.
\begin{asparaitem}
	\item For every $x \in \st$: For every $c \in \mcol(x)$, $x$ is the highest element
	in $T(c)$. Furthermore, $c = \rep(c)$.
	\item Consider $x, y \in \st$ such that $y$ was pushed on $x$. There exists $c \in \col(x) \setminus
	\mcol(x)$ such that $x$ is not highest in $T(c)$ but $y$ is highest in $T(c)$.
\end{asparaitem}

@@

\smallskip
\begin{compactenum}
	\item If $K$ is empty, pick arbitrary $c$. Let $x$ be the highest point in $T(c)$.
	Push $x$ onto stack and set $\mcol(x) = \{c\}$.
	\item Let $h$ be head of $\st$.
	\item While $\col(h) \setminus \mcol(h)$ non-empty:
	\begin{compactenum}
		\item Pick some $c \in \col(h) \setminus \mcol(h)$. Replace it by $c' = \rep(c)$.
		\item Obtain highest point $x$ in $T(c')$. 
		\item If $x = h$, put $c'$ in $\mcol(h)$. Otherwise ($x \neq h$), push $x$ onto stack
		and add $c'$ to $\mcol(x)$.
		\item Let $h$ be new head of $\st$.
	\end{compactenum}
\end{compactenum}
\end{minipage}}

@@

\smallskip
\begin{compactenum}
	\item Run $\update(\st)$. Pop $\st$ to get $h$.
	\item Suppose $\mcol(h) = \{c\}$ (so $h$ is split). 
	\begin{compactenum}
		\item Connect $h$ (in $\reeb(\MM)$) to $\h(c)$ and the unique minimum corresponding to split $h$. 
		\item Delete $h$ from $T(c)$, set $\h(c) = h$.
		\item End procedure.
	\end{compactenum}
	\item Let $\mcol(h) = \{c_1, c_2\}$ (so $h$ is merge).
	\item Connect $h$ in $\reeb(\MM)$ to $\h(c_1)$ and $\h(c_2)$.
	\item Delete all copies of $h$ from $T(c_1)$ and $T(c_2)$.
	\item Perform union of colors $c_1$ and $c_2$ (denote merged color as $c$). Merge heaps
	to get $T(c) = T(c_1) \cup T(c_2)$.
	\item Set $\h(c) = h$.	
\end{compactenum}
\end{minipage}}
